---
title: "DataAnalysis"
author: Elizabeth Orozco & Alex Stadlinger
format: pdf
editor: visual
---

## *Load in packages and unzip files*

```{r}
#load in required packages
library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(brms)
library(ggeffects)
```

```{r}
#unzip rodent activity file
# unzip("edi.2200.1.zip")
```

## *Read in required files*

```{r}
#read in rodent activity file
rodent_data <- read.csv("Rodent_activity.csv")
```

```{r}
#read in trapping data file
trapping_data <- read.csv("Trapping.csv")
```

```{r}
#read in vegetation cover data file
vegetation_data <- read.csv("Vegetation_cover.csv")
```

```{r}
#read in seed predation data
seed_predation_data <- read.csv("Seed_predation.csv")
```

## *Clean up files*

```{r}
#separate data in trapping_data
colnames(trapping_data)

trapping_separated <- trapping_data %>%
  separate(col = Trap_day.Lynx.Habitat.Plot.Trap.Species.Working, 
           into = c("Trap_day", "Lynx", "Habitat", "Plot", "Trap", "Species", "Working"), 
           sep = ";")
```

```{r}
#separate data in vegetation_data
colnames(vegetation_data)

vegetation_separated <- vegetation_data %>%
  separate(col = Lynx.Plot.Habitat.Tree.Shrub.Rock, 
           into = c("Lynx", "Plot", "Habitat", "Tree", "Shrub", "Rock"), 
           sep = ";")
```

```{r}
#separate seed predation data
colnames(seed_predation_data)

seed_predation_separated <- seed_predation_data %>%
  separate(col = Lynx.Habitat.Plot.Microhabitat.Code.CAM.Seeds_left0.Seeds_left1.Seeds_predated1.Seeds_left2.Seeds_oredated2.Seeds_left3.Seeds_predated3.Seeds_left4.Seeds_predated4.Seeds_left5.Seeds_predated5.Seeds_predated_total, 
           into = c("Lynx", "Habitat", "Plot", "Microhabitat", "Code", "CAM", "Seeds_left0", "Seeds_left1", "Seeds_predated1", "Seeds_left2", "Seeds_oredated2", "Seeds_left3", "Seeds_predated3", "Seeds_left4", "Seeds_predated4", "Seeds_left5", "Seeds_predated5", "Seeds_predated_total"), 
           sep = ";")
```

```{r}
#make seed predated total column numeric
seed_predation_separated$Seeds_predated_total <- as.numeric(seed_predation_separated$Seeds_predated_total)

seed_predation_separated <- seed_predation_separated %>% 
  mutate(Seeds_predated_total = as.numeric(Seeds_predated_total),
         Seeds_left0 = as.numeric(Seeds_left0))

```

## *Graphs!*

```{r}
#plotting seed predation total against lynx presence
ggplot(data = seed_predation_separated, aes(x = Lynx, y = Seeds_predated_total)) +
  geom_col()
```

```{r}
#plotting seed predation total as a function of habitat type
ggplot(data = seed_predation_separated, aes(x = Habitat, y = Seeds_predated_total)) +
  geom_col()
```

```{r}
#plotting seed predation total as a function of microhabitat
ggplot(data = seed_predation_separated, aes(x = Microhabitat, y = Seeds_predated_total)) +
  geom_col()
```

```{r}
#graph 1: lynx = 1, microhabitat = open
lynx_open <- filter(seed_predation_separated, Lynx == "1", Microhabitat == "Open")

g1 <- ggplot(data = lynx_open, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) + 
  geom_col() +
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 2: lynx = 1, microhabitat = rock
lynx_rock <- filter(seed_predation_separated, Lynx == "1", Microhabitat == "Rock")

g2 <- ggplot(data = lynx_rock, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 3: lynx = 1, microhabitat = shrub
lynx_shrub <- filter(seed_predation_separated, Lynx == "1", Microhabitat == "Shrub")

g3 <- ggplot(data = lynx_shrub, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 4: lynx = 0, microhabitat = open
noLynx_open <- filter(seed_predation_separated, Lynx == "0", Microhabitat == "Open")

g4 <- ggplot(data = lynx_open, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 5: lynx = 0, microhabitat = rock
noLynx_rock <- filter(seed_predation_separated, Lynx == "0", Microhabitat == "Rock")

g5 <- ggplot(data = noLynx_rock, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 6: lynx = 0, microhabitat = shrub
noLynx_shrub <- filter(seed_predation_separated, Lynx == "0", Microhabitat == "Shrub")

g6 <- ggplot(data = noLynx_shrub, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") + 
  ylim(0, 50)
```

```{r}
#open
(g1 + g4) + plot_annotation(
  title = 'Microhabitat: Open')
```

```{r}
#rock
(g2 + g5) + plot_annotation(
  title = 'Microhabitat: Rock')
```

```{r}
#shrub
(g3 + g6) + plot_annotation(
  title = 'Microhabitat: Shrub')
```

```{r}
lynx_absent <- filter(seed_predation_separated, Lynx == "0")

ggplot(data = lynx_absent, aes(x = Microhabitat, y = Seeds_predated_total, fill = Microhabitat)) +
  geom_col() + 
  labs(x = "Microhabitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") +
  ylim(0, 100) +
  theme_classic()
```

```{r}
lynx_present <- filter(seed_predation_separated, Lynx == "1")

ggplot(data = lynx_present, aes(x = Microhabitat, y = Seeds_predated_total, fill = Microhabitat)) +
  geom_col() + 
  labs(x = "Microhabitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 100) + 
  theme_classic()
```

```{r}
seed_predation_separated %>%

ggplot(aes(x = Lynx, y = Seeds_predated_total, color = Lynx)) +
  geom_jitter(width = 0.1) +
  # labs(x = "Microhabitat",
  #      y = "Total Number of Seeds Predated",
  #      title = "Lynx Present") +
  facet_grid(Habitat ~ Microhabitat) +
  theme_bw() 
```

```{r}
summary_stats <- seed_predation_separated %>%
  group_by(Lynx, Habitat, Microhabitat) %>% 
  summarise(
   mean_seeds = mean(Seeds_predated_total, na.rm = TRUE),
    n = sum(!is.na(Seeds_predated_total)),
    sd_seeds = sd(Seeds_predated_total, na.rm = TRUE),
    se_seeds = sd_seeds / sqrt(n),
    ci_lower = mean_seeds - qt(0.975, df = n - 1) * se_seeds,
    ci_upper = mean_seeds + qt(0.975, df = n - 1) * se_seeds
  ) %>%
  ungroup()

ggplot(summary_stats, aes(x = Lynx, y = mean_seeds, color = Lynx)) +
  geom_point(size = 3) +  # Average points
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # 95% CI error bars
  facet_grid(Habitat ~ Microhabitat) +
  labs(x = "Lynx Presence", y = "Average Number of Seeds Predated", title = "Seed Predation Rates as a Function of Lynx Presence and Micro/Macro habitat") +
  scale_color_manual(values = c("Present" = "goldenrod", "Absent" = "coral"))
```

## *Model*

```{r}
seed_predation_separated %>% 
  count(Lynx, Habitat, Microhabitat, Plot)

# number of seeds predated ~ Microhabitat + Lynx Presence
m.seeds.habitat.microhabitat.lynx <-
brm(data = seed_predation_separated, # Give the model the data
# Choose a poisson distribution - THIS IS THE NEW PART!
family = binomial(link = "logit"),
# Specify the model here.
Seeds_predated_total | trials(Seeds_left0) ~ 1 + Habitat + Microhabitat + Lynx + (1|Plot),
# Here's where you specify parameters for executing the Markov chains
# We're using similar to the defaults, except we set cores to 4 so the analysis runs fa
iter = 2000, warmup = 1000, chains = 4, cores = 4,
# Save the fitted model object as output - helpful for reloading in the output later
file = "output/m.seeds.habitat.microhabitat.lynx")
```

```{r}
summary(m.seeds.habitat.microhabitat.lynx)
```

```{r}
plot(m.seeds.habitat.microhabitat.lynx)
```

```{r}
# number of seeds predated ~ Microhabitat + Lynx Presence
m.seeds.habitat.microhabitat.lynx.int <-
brm(data = seed_predation_separated, # Give the model the data
# Choose a poisson distribution - THIS IS THE NEW PART!
family = binomial(link = "logit"),
# Specify the model here.
Seeds_predated_total | trials(Seeds_left0) ~ 0 + Habitat + Lynx*Microhabitat + (1|Plot),
# Here's where you specify parameters for executing the Markov chains
# We're using similar to the defaults, except we set cores to 4 so the analysis runs fa
iter = 2000, warmup = 1000, chains = 4, cores = 4,
# Save the fitted model object as output - helpful for reloading in the output later
file = "output/m.seeds.habitat.microhabitat.lynx.int")
```

```{r}
summary(m.seeds.habitat.microhabitat.lynx.int)
```

## Model Summary

### Additive Model:

Intercept = Forest Habitat, Open Microhabitat, No Lynx

Open habitat: 0.16 increase in seeds predated compared to forest (not reasonably different from 0)

Rock Microhabitat: 0.94 increase in seeds predated compared to open (reasonably different from 0)

Shrub Microhabitat: 2.19 increase in seeds predated compared to open (reasonably different from 0)

Lynx Present: 0.92 increase in seeds predated compared to lynx absent (not reasonably different from 0)

```{r}
loo(m.seeds.habitat.microhabitat.lynx)
```

```{r}
waic(m.seeds.habitat.microhabitat.lynx)
```

### Interactive Model:

Difference from 0

Forest (-2.91) and open (-2.77) habitats are reasonably different from 0.

In rock microhabitat, presence of lynx is not reasonably different from 0. This means that presence or absence of lynx does not significantly alter seed predation success in rock microhabitat.

In shrub microhabitat, presence of lynx is not reasonably different from 0. This means that presence or absence of lynx does not significantly alter seed predation success in shrub microhabitat. Approaching on 0, so there is more of a difference between predator presence or absence than in rock microhabitat.

```{r}
loo(m.seeds.habitat.microhabitat.lynx.int)
```

```{r}
waic(m.seeds.habitat.microhabitat.lynx.int)
```

### Predictive Plot

```{r}
pred.add <- predict_response(m.seeds.habitat.microhabitat.lynx, condition = c(Seeds_left0 = 8))
plot(pred)
```

### Compare Models

The additive model is better because it has lower PSIS (731.7) and WAIC (719.9) values compared to the interactive model's PSIS (751.6) and WAIC (739.3) values.
