---
title: "Data Exploration"
author: Elizabeth Orozco & Alex Stadlinger
format: pdf
editor: visual
---

# *Load in Packages*

```{r}
#load in required packages
library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(brms)
library(ggeffects)
```

# Unzip and Read in Files

## Unzip

```{r}
#unzip rodent activity file
# unzip("edi.2200.1.zip")
```

## *Read in files*

```{r}
#read in rodent activity file
rodent_data <- read.csv("data/Rodent_activity.csv")
```

```{r}
#read in trapping data file
trapping_data <- read.csv("data/Trapping.csv")
```

```{r}
#read in vegetation cover data file
vegetation_data <- read.csv("data/Vegetation_cover.csv")
```

```{r}
#read in seed predation data
seed_predation_data <- read.csv("data/Seed_predation.csv")
```

# *Data Wrangling*

## Separate 

```{r}
#separate data in trapping_data
colnames(trapping_data) #obtain column names to know what to call new columns

#use separate to separate on '.'
trapping_separated <- trapping_data %>%
  separate(col = Trap_day.Lynx.Habitat.Plot.Trap.Species.Working, 
           into = c("Trap_day", "Lynx", "Habitat", "Plot", "Trap", "Species", "Working"), 
           sep = ";")
```

```{r}
#separate data in vegetation_data
colnames(vegetation_data)

vegetation_separated <- vegetation_data %>%
  separate(col = Lynx.Plot.Habitat.Tree.Shrub.Rock, 
           into = c("Lynx", "Plot", "Habitat", "Tree", "Shrub", "Rock"), 
           sep = ";")
```

```{r}
#separate seed predation data
colnames(seed_predation_data)

seed_predation_separated <- seed_predation_data %>%
  separate(col = Lynx.Habitat.Plot.Microhabitat.Code.CAM.Seeds_left0.Seeds_left1.Seeds_predated1.Seeds_left2.Seeds_oredated2.Seeds_left3.Seeds_predated3.Seeds_left4.Seeds_predated4.Seeds_left5.Seeds_predated5.Seeds_predated_total, 
           into = c("Lynx", "Habitat", "Plot", "Microhabitat", "Code", "CAM", "Seeds_left0", "Seeds_left1", "Seeds_predated1", "Seeds_left2", "Seeds_oredated2", "Seeds_left3", "Seeds_predated3", "Seeds_left4", "Seeds_predated4", "Seeds_left5", "Seeds_predated5", "Seeds_predated_total"), 
           sep = ";")
```

## As.Numeric

```{r}
#make total seeds predated column and seeds left 0 column numeric
#Seeds_left0 is used as total seeds left out at the beginning (8 seeds)

seed_predation_separated <- seed_predation_separated %>% 
  mutate(Seeds_predated_total = as.numeric(Seeds_predated_total),
         Seeds_left0 = as.numeric(Seeds_left0))

```

# *Graphs!*

## 3 Simple Bar Graphs

```{r}
#plotting seed predation total as a function of lynx presence
ggplot(data = seed_predation_separated, aes(x = Lynx, y = Seeds_predated_total)) +
  geom_col()

ggsave("output/SeedPredationLynxPresence.png") #save to output folder
```

```{r}
#plotting seed predation total as a function of macrohabitat type
ggplot(data = seed_predation_separated, aes(x = Habitat, y = Seeds_predated_total)) +
  geom_col()

ggsave("output/SeedPredationMacrohabitat.png")
```

```{r}
#plotting seed predation total as a function of microhabitat type
ggplot(data = seed_predation_separated, aes(x = Microhabitat, y = Seeds_predated_total)) +
  geom_col()

ggsave("output/SeedPredationMicrohabitat.png")
```

## 6 Bar Graphs with All Variables

```{r}
#graph 1: lynx = 1, microhabitat = open

#make new data frame with just rows where lynx is present AND microhabitat is open
lynx_open <- filter(seed_predation_separated, Lynx == "1", Microhabitat == "Open")

g1 <- ggplot(data = lynx_open, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) + 
  geom_col() +
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 2: lynx = 1, microhabitat = rock

#make new data frame with just rows where lynx is present AND microhabitat is rock
lynx_rock <- filter(seed_predation_separated, Lynx == "1", Microhabitat == "Rock")

g2 <- ggplot(data = lynx_rock, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 3: lynx = 1, microhabitat = shrub

#make new data frame with just rows where lynx is present AND microhabitat is shrub
lynx_shrub <- filter(seed_predation_separated, Lynx == "1", Microhabitat == "Shrub")

g3 <- ggplot(data = lynx_shrub, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 4: lynx = 0, microhabitat = open

#make new data frame with just rows where lynx is not present AND microhabitat is open
noLynx_open <- filter(seed_predation_separated, Lynx == "0", Microhabitat == "Open")

g4 <- ggplot(data = lynx_open, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 5: lynx = 0, microhabitat = rock

#make new data frame with just rows where lynx is not present AND microhabitat is rock
noLynx_rock <- filter(seed_predation_separated, Lynx == "0", Microhabitat == "Rock")

g5 <- ggplot(data = noLynx_rock, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") + 
  ylim(0, 50)
```

```{r}
#graph 6: lynx = 0, microhabitat = shrub

#make new data frame with just rows where lynx is not present AND microhabitat is shrub
noLynx_shrub <- filter(seed_predation_separated, Lynx == "0", Microhabitat == "Shrub")

g6 <- ggplot(data = noLynx_shrub, aes(x = Habitat, y = Seeds_predated_total, fill = Habitat)) +
  geom_col() + 
  labs(x = "Habitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") + 
  ylim(0, 50)
```

```{r}
#put the two graphs with the open microhabitat type together
(g1 + g4) + plot_annotation(
  title = 'Microhabitat: Open')

ggsave("output/SeedPredationLynxPresenceMacrohabitatMicrohabitatOpen.png")
```

```{r}
##put the two graphs with the rock microhabitat type together
(g2 + g5) + plot_annotation(
  title = 'Microhabitat: Rock')

ggsave("output/SeedPredationLynxPresenceMacrohabitatMicrohabitatRock.png")
```

```{r}
#put the two graphs with the shrub microhabitat type together
(g3 + g6) + plot_annotation(
  title = 'Microhabitat: Shrub')

ggsave("output/SeedPredationLynxPresenceMacrohabitatMicrohabitatShrub.png")
```

## 2 Bar Graphs of Seeds Predated Against Microhabitat Type

```{r}
#create a new data frame where all rows have no lynx present
lynx_absent <- filter(seed_predation_separated, Lynx == "0")

#graph seeds predated as a function of microhabitat type
ggplot(data = lynx_absent, aes(x = Microhabitat, y = Seeds_predated_total, fill = Microhabitat)) +
  geom_col() + 
  labs(x = "Microhabitat",
       y = "Total Number of Seeds Predated",
       title = "No Lynx Present") +
  ylim(0, 100) +
  theme_classic()

ggsave("output/SeedPredationNoLynxMicrohabitat.png")
```

```{r}
#create a new data frame where all rows have lynx present
lynx_present <- filter(seed_predation_separated, Lynx == "1")

#graph seeds predated as a function of microhabitat type
ggplot(data = lynx_present, aes(x = Microhabitat, y = Seeds_predated_total, fill = Microhabitat)) +
  geom_col() + 
  labs(x = "Microhabitat",
       y = "Total Number of Seeds Predated",
       title = "Lynx Present") + 
  ylim(0, 100) + 
  theme_classic()

ggsave("output/SeedPredationLynxPresentMicrohabitat.png")
```

## Jitter Plot with all 4 Variables Represented 

```{r}
#Graph seeds predated in 6 panels representing the different combinations of experimental conditions. Each panel has lynx/no lynx and combines all 6 combinations of macro- and microhabitat.

seed_predation_separated %>%
ggplot(aes(x = Lynx, y = Seeds_predated_total, color = Lynx)) +
  geom_jitter(width = 0.1) +
  # labs(x = "Microhabitat",
  #      y = "Total Number of Seeds Predated",
  #      title = "Lynx Present") +
  facet_grid(Habitat ~ Microhabitat) +
  theme_bw() 

ggsave("output/SeedPredationLynxMacrohabitatMicrohabitat.png")
```

## Final Visualization

```{r}
summary_stats <- seed_predation_separated %>%
  group_by(Lynx, Habitat, Microhabitat) %>% 
  summarise(
   mean_seeds = mean(Seeds_predated_total, na.rm = TRUE),
    n = sum(!is.na(Seeds_predated_total)),
    sd_seeds = sd(Seeds_predated_total, na.rm = TRUE),
    se_seeds = sd_seeds / sqrt(n),
    ci_lower = mean_seeds - qt(0.975, df = n - 1) * se_seeds,
    ci_upper = mean_seeds + qt(0.975, df = n - 1) * se_seeds
  ) %>%
  ungroup()

ggplot(summary_stats, aes(x = Lynx, y = mean_seeds, color = Lynx)) +
  geom_point(size = 3) +  # Average points
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +  # 95% CI error bars
  facet_grid(Habitat ~ Microhabitat) +
  labs(x = "Lynx Presence", y = "Average Number of Seeds Predated", title = "Seed Predation Rates as a Function of Lynx Presence and Micro/Macro habitat") +
  scale_color_manual(values = c("Present" = "goldenrod", "Absent" = "coral"))

ggsave("output/FinalVisualization.png")
```

# *Models*

## Additive Model

```{r}
#Count the number of trials for each combination of experimental variables (just to see how many trials at each plot)
seed_predation_separated %>% 
  count(Lynx, Habitat, Microhabitat, Plot)

#Run an addtive model of number of seeds predated ~ Microhabitat + Lynx Presence
m.seeds.habitat.microhabitat.lynx <-
brm(data = seed_predation_separated, # Give the model the data
# Choose a binomial distribution --> There is a set number (8 seeds) and it is success vs. failure
family = binomial(link = "logit"),
# Specify the model here.
Seeds_predated_total | trials(Seeds_left0) ~ 1 + Habitat + Microhabitat + Lynx + (1|Plot),
# Specify parameters for executing the Markov chains
iter = 2000, warmup = 1000, chains = 4, cores = 4,
# Save the model in output folder
file = "output/m.seeds.habitat.microhabitat.lynx")
```

```{r}
#get summary of the additive model
summary(m.seeds.habitat.microhabitat.lynx)
```

```{r}
#check that model ran correctly
plot(m.seeds.habitat.microhabitat.lynx)
```

## Interactive Model

```{r}
# Run an interactive model of number of seeds predated ~ Microhabitat + Lynx Presence
m.seeds.habitat.microhabitat.lynx.int <-
brm(data = seed_predation_separated, # Give the model the data
family = binomial(link = "logit"),
# Specify the model here.
Seeds_predated_total | trials(Seeds_left0) ~ 0 + Habitat + Lynx*Microhabitat + (1|Plot),
# Specify parameters for executing the Markov chains
iter = 2000, warmup = 1000, chains = 4, cores = 4,
file = "output/m.seeds.habitat.microhabitat.lynx.int")
```

```{r}
#get summary of the interactive model
summary(m.seeds.habitat.microhabitat.lynx.int)
```

```{r}
#check that model ran correctly
plot(m.seeds.habitat.microhabitat.lynx.int)
```

# Model Summary

## Additive Model:

Intercept = Forest Habitat, Open Microhabitat, No Lynx

Open habitat: 0.16 increase in seeds predated compared to forest (not reasonably different from 0)

Rock Microhabitat: 0.94 increase in seeds predated compared to open (reasonably different from 0)

Shrub Microhabitat: 2.19 increase in seeds predated compared to open (reasonably different from 0)

Lynx Present: 0.92 increase in seeds predated compared to lynx absent (not reasonably different from 0)

```{r}
#check PSIS value
loo(m.seeds.habitat.microhabitat.lynx)
```

```{r}
#check WAIC value
waic(m.seeds.habitat.microhabitat.lynx)
```

## Interactive Model:

Difference from 0

Forest (-2.91) and open (-2.77) habitats are reasonably different from 0.

In rock microhabitat, presence of lynx is not reasonably different from 0. This means that presence or absence of lynx does not significantly alter seed predation success in rock microhabitat.

In shrub microhabitat, presence of lynx is not reasonably different from 0. This means that presence or absence of lynx does not significantly alter seed predation success in shrub microhabitat. Approaching on 0, so there is more of a difference between predator presence or absence than in rock microhabitat.

```{r}
#check PSIS value
loo(m.seeds.habitat.microhabitat.lynx.int)
```

```{r}
#check WAIC value
waic(m.seeds.habitat.microhabitat.lynx.int)
```

# Predictive Plots

```{r}
# run a predictive plot to see help evaluate model correctness
pred.add <- predict_response(m.seeds.habitat.microhabitat.lynx, condition = c(Seeds_left0 = 8))
plot(pred)
```

# Compare Models

The additive model is better because it has lower PSIS (731.7) and WAIC (719.9) values compared to the interactive model's PSIS (751.6) and WAIC (739.3) values.
